---
title: "GAMMs Session 3"
subtitle: "Model Comparison and Autoregression"
author: "Joshua <span style='font-variant: small-caps;'>Wilson Black</span>"
institute: "Te K캐hui Roro Reo | New Zealand Institute of Language, Brain and Behaviour <br/> Te Whare W캐nanga o Waitaha | University of Canterbury"
format:
  revealjs:
    theme: [custom.scss]
    incremental: true
    logo: images/NZILBB-small.svg
    df-print: paged
    title-slide-attributes:
      data-background-image: ../images/NZILBB-big.svg
      data-background-position: 'top left'
      data-background-size: 50%
bibliography: 
  - grateful-refs.bib
  - ../../stat_workshops.bib
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
---

```{r}
#| echo: false
library(tidyverse)
library(here)
library(parallel)
library(mgcv)
library(itsadug)

theme_set(theme_bw())

# Load last weeks data behind the scenes
price <- read_rds(here('data', 'price_anon.rds'))

price_filtered <- price |> 
  filter(
    between(f1, f1_lower, f1_upper),
    between(f2, f2_lower, f2_upper),
    syllables_per_sec < 10,
    following_voiceless == FALSE # filtering step used in paper.
  ) |> 
  mutate(
    sex = factor(sex),
    speaker_anon = factor(speaker_anon),
  ) |> 
  # create identifier for each trajectory.
  group_by(speaker_anon) |> 
  mutate(
    traj_no = cumsum(measurement_no <= lag(measurement_no, default = 11)),
    traj_id = paste0(speaker_anon, '_', traj_no),
    traj_id = factor(traj_id)
  ) |> 
  ungroup()

# make a subset of the data so that it can be fit on a laptop. Just keep 
# the first three trajectories and 30% of the speakers. We want some early
# and some later speakers.
early_speakers <- price |> 
  filter(yob < 1900) |> 
  pull(speaker_anon) |>
  unique()

mid_speakers <- price |> 
  filter(between(yob, 1900, 1950)) |> 
  pull(speaker_anon) |>
  unique()

late_speakers <- price |> 
  filter(yob > 1950) |> 
  pull(speaker_anon) |>
  unique()

speakers_to_keep <- c(
  sample(early_speakers, size = 50),
  sample(mid_speakers, size = 50),
  sample(late_speakers, size = 50)
)

price_filtered <- price_filtered |> 
  filter(
    speaker_anon %in% speakers_to_keep,
    traj_no <= 3
  )
```


## Overview

1. Recap
2. Autocorrelation
3. Model comparison
   
# Recap

## What are GAMMs?

- A flexible, but still interpretable, way of capturing non-linear patterns
in data.
- [G]{.red}eneralised: use variety of response families (just like GLMMs).
- [A]{.red}dditive: sum together smooth functions of our explanatory variables.
- [M]{.red}ixed: include random effects (incl. random smooths)
- [M]{.red}odel: ...they're statistical models.

## `mgcv` Syntax

::: {.fragment}

```{r}
#| cache: true
#| code-line-numbers: "|1|2|3-4,6|5|7|8|10-11"
f1_fs_fit <- bam(
  formula = f1 ~ sex + 
    s(measurement_no, by = sex, k = 4, bs = "cr") +
    s(yob, by = sex, k = 4, bs = "cr") + 
    ti(measurement_no, yob, by = sex, k = c(4, 4), bs = "cr") +
    s(syllables_per_sec, k = 4, bs = "cr") + 
    s(speaker_anon, bs = "re") +
    s(measurement_no, traj_id, bs = "fs", k = 4, m = 1),
  data = price_filtered,
  discrete = TRUE,
  nthreads = detectCores() - 1
)
```

:::

::: notes

This is the most complex model from our previous session. We're looking at 
F1 trajectories for [price]{.smallcaps}.

1. `mgcv` functions: `gam`, `gamm`, `bam`. Latter for large data sets. First,
for gams witout random effects structures.
2. Explaining `f1` via, a parametric term sex with `M` and `F`.
3. A series of smooth terms of one continuous variable. Measurement number,
i.e., time within a single diphthong token. We've specified k, in this
case with some knowledge from previous models that we've got enough wiggliness
and with a view to decreasing the computational load.
4. A continuous interaction term which captures the interaction between
measurement number and year of birth.
5. For both (3) and (4), we are fitting for both male and female speakers.
6. A by-speaker random intercept.
7. Random smooths for each trajectory. Note `m=1`.
8. Use of `discrete=TRUE` and multiple cores (provided with `bam`) to speed
up model fit.

:::

## The Story So Far

- What GAMMs are.
- How to fit a GAM with `mgcv`.
  - Some common error messages
  - Difference smooths vs. by-factor smooths
- How to fit a GA[M]{.red}M with `mgcv`
  - Random intercepts,
  - Random slopes, 
  - Random smooths
  
# Autocorrelation

## Independent Information {.smaller}

- A measurement at time $t$ is often closely correlated with the measurement
at time $t-1$.
    - If I have a thermometer in my office, the measurement at 1:00pm and the
    measure at 1:01pm are going to be very similar.
    - Four measures
    spread across different times and days carry more information than 
    measures taken on the same day at: 1:00pm, 1:01pm, 1:02pm, 1:03pm.
- This kind of dependence is called [autocorrelation]{.red}.
- You [must]{.red} think about this when using GA(M)Ms.

## Testing for Autocorrelation

- We use the function `acf()` from `itsadug`.
- Let's create a bad model.

## Ignoring Autocorrelation

```{r}
bad_fit <- bam(
  formula = f1 ~ sex + 
    s(measurement_no, by = sex, k = 4, bs = "cr") +
    s(yob, by = sex, k = 4, bs = "cr") + 
    ti(measurement_no, yob, by = sex, k = c(4, 4), bs = "cr") +
    s(syllables_per_sec, k = 4, bs = "cr") + 
    s(speaker_anon, bs = "re"),
  data = price_filtered,
  discrete = TRUE,
  nthreads = detectCores() - 1
)
```

## Apply `acf()`

::: fragment
```{r}
#| output-location: fragment
acf_resid(bad_fit)
```
:::

- At lag=1, autocorrelation gets up to around 0.5.


## What to Do?

- Random Smooths
    - Computational overkill (or even failure!)
    - Especially when using `method="ML"` (see model comparisons below).
- AR(1) autocorrelation model
    - Adds a dependence on the previous observation
    - Less computationally intensive.
    
## Random Smooths?

::: fragment
```{r}
#| output-location: fragment
acf_resid(f1_fs_fit, split_pred="traj_id")
```
:::

- Some autocorrelation left at lags 1 through 4. Largest is around -0.2, not 
a big deal!
- Often not an issue as $R^2$ is so high: `r round(summary(f1_fs_fit, re.test = FALSE)$r.sq * 100)`%

## Trying an AR1 model

- We have to work out $\rho$.
- **Preprocessing:** mark the start of each trajectory.
    - We need a new column with `TRUE` at the first entry
    of each trajectory.
    - **NB:** subsequent rows must be subsequent observations.
    
---

```{r}
price_filtered <- price_filtered |> 
  mutate(
    traj_start = traj_id != lag(
      traj_id, 
      # 'default' must be something other than the first traj_id.
      default=price_filtered$traj_id[12]
    )
  )
```

---

```{r}
#| eval: false
# An alternative: use the start_event function in `itsadug`
price_filtered <- start_event(
  as.data.frame(price_filtered), # This doesn't work with tibbles.
  column = "measurement_no", # the column containing relative time information..
  event="traj_id"
)
```

## Work out $\rho$

::: fragment
```{r}
#| output-location: fragment
r1 <- start_value_rho(bad_fit, plot=TRUE) # plot=FALSE suppressed the plot.
```
:::

## Apply AR1

```{r}
ar1_fit <- bam(
  formula = f1 ~ sex + 
    s(measurement_no, by = sex, k = 4, bs = "cr") +
    s(yob, by = sex, k = 4, bs = "cr") + 
    ti(measurement_no, yob, by = sex, k = c(4, 4), bs = "cr") +
    s(syllables_per_sec, k = 4, bs = "cr") + 
    s(speaker_anon, bs = "re"),
  data = price_filtered,
  rho = r1,
  AR.start = price_filtered$traj_start,
  discrete = TRUE,
  nthreads = detectCores() - 1
)
```

## Did it Work?

::: fragment
```{r}
#| output-location: fragment
acf_resid(ar1_fit, split_pred="AR.start")
```
:::

- Certainly at lag 1, with some minor residual issues at lag 2 and 3.

# Model Comparison

## Motivating Questions

- Is it worth including that interaction?
- What is the overall influence of $x$ on the model?
    - Esp. where $x$ appears in multiple places?
    - What if my hypothesis doesn't correspond to a particular coefficient
    in the summary?
- How can I choose between multiple, theoretically defensible, models?

## What is Model Comparison? {.smaller}

- We use various measures of model fit to compare models.
- A bad choice: $R^2$ --- it always increases with more predictors.
- Better: a measure which penalises complexity.
    - Is the addition of _complexity_ worth it for the increase in _prediction_.
- Often, we test nested models.
    - i.e. a bigger model includes all of the predictors in the
    smaller model.
- What about backwards and forward selection?
    - Not my preference and GAMMs make the computational costs even worse.
    - Include the stuff you think might be relevant!

## Using `compareML()`

- `itsadug` has a useful function called `compareML()` which runs a
$\chi^2$ test tailored for GAMMs.
- `AIC()` is another option but can't be used with AR1 autocorrelation models (more later!)
- Simply put the two models in as arguments.
- **Spiky bit:** You have to use `method="ML"` in `bam()` if your fixed 
effects differ.
    - This _massively_ increases the time required to fit the model.

## `compareML()` Example

- Let's work out if including `sex` is important for our model.
- `sex` appears in **four** places. Let's remove them all and fit a 
new model.

## Full Model {auto-animate=true}

```{r}
#| cache: true
#| code-line-numbers: "|2-5"
ar1_ml_fit <- bam(
  formula = f1 ~ sex + 
    s(measurement_no, by = sex, k = 4, bs = "cr") +
    s(yob, by = sex, k = 4, bs = "cr") + 
    ti(measurement_no, yob, by = sex, k = c(4, 4), bs = "cr") +
    s(syllables_per_sec, k = 4, bs = "cr") + 
    s(speaker_anon, bs = "re"),
  data = price_filtered,
  rho = r1,
  AR.start = price_filtered$traj_start,
  method = "ML"
)
```


## Nested Model {auto-animate=true}

```{r}
#| cache: true
#| code-line-numbers: "2-5"
ar1_nosex_fit <- bam(
  formula = f1 ~ 
    s(measurement_no, k = 4, bs = "cr") +
    s(yob, k = 4, bs = "cr") + 
    ti(measurement_no, yob, k = c(4, 4), bs = "cr") +
    s(syllables_per_sec, k = 4, bs = "cr") + 
    s(speaker_anon, bs = "re"),
  data = price_filtered,
  rho = r1,
  AR.start = price_filtered$traj_start,
  method = "ML"
)
```


## Apply `compareML()`

```{r}
compareML(ar1_nosex_fit, ar1_ml_fit)
```


## Apply `compareML()` (cont.)

- The inclusion of `sex` is statistically significant.
- We can report the results of this test in a paper.
- Note: the order of the two models in `compareML()` doesn't matter.
    - i.e. `compareML(fit_1, fit_2)` is the same as `compareML(fit_2, fit_1)`.
- **Exercice:** Determine $\rho$ for the no-sex model.

## Significance Testing {.smaller}

::: fragment
We've now seen two methods for significance testing:
:::

1. Reading off model summaries.
    - Work with `fREML` (and thus `bam` parallel computation)
    - But is your hypothesis captured in any particular coefficient?
2. ANOVA via `compareML()`
    - More flexible.
    - Can take a long time!

::: fragment
We should also consider:
:::

3. Visual methods

---

```{r}
#| echo: false
#| warning: false
#| message: false
price_offset_means <- price |>
  filter(
    measurement_no == 9,
    between(f1, f1_lower, f1_upper),
    between(f2, f2_lower, f2_upper),
    following_voiceless == FALSE 
  ) |> 
  group_by(speaker_anon) |> 
  summarise(
    f1 = mean(f1, na.rm=TRUE),
    f2 = mean(f2, na.rm=TRUE),
    yob = first(yob),
    sex = first(sex),
    syllables_per_sec = mean(
      syllables_per_sec, 
      na.rm=TRUE
    )
  ) |> 
  mutate(
    sex = ordered(sex)
  )

contrasts(price_offset_means$sex) <- "contr.treatment"

f1_fit <- gam(
  formula = f1 ~ sex + 
    s(yob, k=10, bs = "cr") + 
    s(yob, by = sex, k = 10, bs = "cr") + 
    s(syllables_per_sec, k = 10, bs = "cr"),
  data = price_offset_means
)

plot_diff(
  f1_fit, 
  view = "yob", 
  comp=list(sex=c("M","F")), 
  rm.ranef=T,
  print.summary = FALSE
)
```

## Visual significance testing

- Smooths across factors.
- Bad idea: claim significance if zero falls out the confidence interval of the
_difference smooth_.

::: fragment
::: callout-important
This has an **insane** false positive rate!
:::
:::

## Visual significance testing (cont.)

- M치rton S칩skuthy's advice:
    1. Do model comparison with `compareML()` to determine _overall significance_.
    2. **Then** use a difference smooth to determine *where* the differences are.
- Alternative: make an hypothesis about exactly where you will see a difference
before looking.
  - e.g., we predict the difference smooth will exclude 0 from 20%-60%.
  
# What now?

::: {.fragment}

```{r}
#| eval: false
#install.packages('usethis')
usethis::use_course('nzilbb/gamm_sessions')
```

:::

- A new R project should open in R Studio.
- Go to the `session_3.R` script in the `scripts folder`.
- Work through the script.


# References

```{r}
#| echo: false
grateful::nocite_references(
  grateful::cite_packages(output = "citekeys", out.dir = here())
)
```


::: refs

:::
