---
title: "GAMMs Session 2"
subtitle: "Interactions and Random Effects"
author: "Joshua <span style='font-variant: small-caps;'>Wilson Black</span>"
institute: "Te Kāhui Roro Reo | New Zealand Institute of Language, Brain and Behaviour <br/> Te Whare Wānanga o Waitaha | University of Canterbury"
format:
  revealjs:
    theme: [custom.scss]
    incremental: true
    logo: images/NZILBB-small.svg
    df-print: paged
    title-slide-attributes:
      data-background-image: ../images/NZILBB-big.svg
      data-background-position: 'top left'
      data-background-size: 50%
bibliography: 
  - grateful-refs.bib
  - ../../stat_workshops.bib
editor: 
  markdown: 
    wrap: 72
execute: 
  echo: true
---

```{r}
#| echo: false
library(tidyverse)
library(here)
library(parallel)

theme_set(theme_bw())
```

## Last Time

1. Conceptual introduction
2. Fit a GAM with `mgcv`
3. Simple model evaluation
4. Some plotting

## This Time

1. Another GAM
2. Adding interactions
2. Mixed effects
    - Intercepts
    - Slopes
    - Smooths
    

## Our Data

- We take data from @soskuthyHorizontalDiphthongShift2019
- Exploratory research on change in two diphthongs of New Zealand English.
- We'll look at one: [price]{.smallcaps}
- What does this mean?

---

{{< video PRICE_male_icphs_vid.mp4 width="800" height="600" >}}

<https://github.com/soskuthy/nz_vowels/blob/master/videos/PRICE_male_icphs_vid.mp4>

## Our Data (Sketched)

- We have a large collection of audio recordings of individual instances (tokens)
of these vowels.
- We are interested in the first and second formants (peaks in the
audio caused by position of tongue in mouth &c. &c. &c.)
- Each is sampled 11 times.
- There's more in the data frame (including the _third_ formant, duration of token).

---

```{r}
#| df-print: paged
price <- read_rds(here('data', 'price_anon.rds'))
price |> 
  select(speaker_anon, sex, yob, syllables_per_sec, measurement_no, f1, f2) |> 
  head(n=12) 
```

## A Simple GAM

- What can we do with _one measurement per speaker_?
- Diphthongs go from one place to another.
    - Look at change in 'onset' or 'offset' from 1957-1987 (year of birth).
    - Just like the GAMs we looked at last time!

## Filter Data

```{r}
price_offset_means <- price |>
  filter(
    measurement_no == 9,
    between(f1, f1_lower, f1_upper),
    between(f2, f2_lower, f2_upper),
    following_voiceless == FALSE 
  ) |> 
  group_by(speaker_anon) |> 
  summarise(
    f1 = mean(f1, na.rm=TRUE),
    f2 = mean(f2, na.rm=TRUE),
    yob = first(yob),
    sex = first(sex),
    syllables_per_sec = mean(
      syllables_per_sec, 
      na.rm=TRUE
    )
  )
```

## Fit GAM

```{r}
#| error: true
library(mgcv)
f1_fit <- bam(
  formula = f1 ~ sex + s(yob, by = sex, k = 10, bs = "tp") + 
    s(syllables_per_sec, k = 10, bs = "tp"),
  data = price_offset_means
)
```
::: {.fragment .big}
😤
:::

- `mgcv` wants [factor]{.red} columns.

## Factor conversion

```{r}
price_offset_means <- price_offset_means |> 
  mutate(
    sex = factor(sex)
  )
```

## Try again!

```{r}
f1_fit <- gam(
  formula = f1 ~ sex + 
    s(yob, by = sex, k = 10, bs = "cr") + 
    s(syllables_per_sec, k = 10, bs = "cr"),
  data = price_offset_means
)
```

## Summary

```{r}
summary(f1_fit)
```

## Is `k` high enough?

```{r}
#| fig.keep: none
gam.check(f1_fit)
```

---

```{r}
library(gratia)
appraise(f1_fit)
```

---

```{r}
draw(f1_fit)
```

---

<!-- It looks like I'm plotting twice. This 
lets me output the text output first then make
a new slide with the plot (hiding the code chunk)-->

```{r}
#| fig-keep: none
library(itsadug)
plot_smooth(
  f1_fit,
  view = "yob",
  plot_all = "sex"
)
```

---

```{r}
#| echo: false
library(itsadug)
plot_smooth(
  f1_fit,
  view = "yob",
  plot_all = "sex",
  print.summary = FALSE
)
```

# Interactions

## Interactions

- You've already seen [continuous by factor]{.red} interactions (i.e. `by`).
- What about [continuous by continuous]{.red}?
    - `ti()` - interaction separated from main effects.
    - `te()` - interaction and main effects together.

## GAM with `ti`

```{r}
f1_ti_fit <- bam(
  formula = f1 ~ sex + s(yob, by = sex, k = 10, bs = "tp") + 
    s(syllables_per_sec, k = 4, bs = "tp") + 
    ti(yob, syllables_per_sec, k = c(10, 4), bs = "tp"),
  data = price_offset_means
)
```

---

```{r}
summary(f1_ti_fit)
```

---

```{r}
draw(f1_ti_fit)
```

---

```{r}
#| output-location: slide
pvisgam(
  f1_ti_fit, 
  view=c("yob", "syllables_per_sec"),
  select = 4
)
```


## GAM with `te`

```{r}
f1_te_fit <- gam(
  formula = f1 ~ sex + 
    te(yob, syllables_per_sec, by = sex, k = c(10, 4), bs = "tp"),
  data = price_offset_means
)
```

---

```{r}
summary(f1_te_fit)
```

---

```{r}
draw(f1_te_fit)
```

# Random Effects

## Random Effects (Revision)

- We have lots of readings from each speaker.
- These are not [independent]{.red}.
- Without random effects, our models assume independence.
- This can lead to overconfidence.

## Random Intercepts

- Capture differences in overall mean value of a speaker.
- Assume speakers come from normal distribution.
- We can check this assumption.

## Filtering Again

```{r}
price_filtered <- price |> 
  filter(
    measurement_no == 9,
    between(f1, f1_lower, f1_upper),
    between(f2, f2_lower, f2_upper),
    following_voiceless == FALSE, # filtering step used in paper.
    syllables_per_sec < 10
  ) |> 
  mutate(
    sex = factor(sex)
  )
```


## GAMM with Random Intercepts

```{r}
#| error: true
f1_int_fit <-  bam(
  formula = f1 ~ sex + s(yob, by = sex, k = 10, bs = "tp") + 
    s(syllables_per_sec, k = 4, bs = "tp") + 
    ti(yob, syllables_per_sec, k = c(10, 4), bs = "tp") +
    s(speaker_anon, bs = "re"),
  data = price_filtered
)
```

::: {.fragment .big}
🤦
:::

- Same problem, different error message!

---

```{r}
price_filtered <- price_filtered |> 
  mutate(
    speaker_anon = factor(speaker_anon)
  )
```

## GAMM with Random Intercepts (2)

```{r}
#| cache: true
f1_int_fit <-  gam(
  formula = f1 ~ sex + s(yob, by = sex, k = 10, bs = "tp") + 
    s(syllables_per_sec, k = 4, bs = "tp") + 
    ti(yob, syllables_per_sec, k = c(10, 4), bs = "tp") +
    s(speaker_anon, bs = "re"),
  data = price_filtered
)
```

- Things are starting to slow down!

---

```{r}
summary(f1_int_fit)
```

---

```{r}
appraise(f1_int_fit)
```

---

```{r}
draw(f1_int_fit)
```

## Random Slopes

- Usually factor x factor
- We'll make another factor:

::: fragment
```{r}
price_filtered <- price_filtered |> 
  mutate(
    following_liquid = following %in% c("r", "l"),
    following_liquid = factor(following_liquid)
  )
```
:::

## GAMM with Random Slopes

```{r}
#| code-line-numbers: "|6|1|3-5|8-9"
#| cache: true
f1_int_slope_fit <-  bam(
  formula = f1 ~ sex + following_liquid + 
    s(yob, by = sex, k = 10, bs = "cr") + 
    s(syllables_per_sec, k = 4, bs = "cr") + 
    ti(yob, syllables_per_sec, k = c(10, 4), bs = "cr") +
    s(speaker_anon, following_liquid, bs = "re"),
  data = price_filtered,
  discrete = TRUE,
  nthreads = detectCores() - 1
)
```

---

```{r}
summary(f1_int_slope_fit)
```

## Random Smooths

- We have multiple _trajectories_ for each speaker.
- We can capture this feature of the full data set
with a factor smooth interaction.
- Size and computational resources are growing rapidly:
  - `f1_ti_fit` < 1 MB
  - `f1_int_fit` = 22.7 MB
  - `f1_int_slope_fit` = 87.6 MB
  - `f1_fs_fit` = 231.9 MB
- **NB**: different data sets.
  
## Filtering (again)

```{r}
price_filtered <- price |> 
  filter(
    between(f1, f1_lower, f1_upper),
    between(f2, f2_lower, f2_upper),
    syllables_per_sec < 10,
    following_voiceless == FALSE # filtering step used in paper.
  ) |> 
  mutate(
    sex = factor(sex),
    speaker_anon = factor(speaker_anon),
  ) |> 
  # create identifier for each trajectory.
  group_by(speaker_anon) |> 
  mutate(
    traj_no = cumsum(measurement_no <= lag(measurement_no, default = 11)),
    traj_id = paste0(speaker_anon, '_', traj_no),
    traj_id = factor(traj_id)
  ) |> 
  ungroup()

# make a subset of the data so that it can be fit on a laptop. Just keep 
# the first three trajectories and 30% of the speakers. We want some early
# and some later speakers.
early_speakers <- price |> 
  filter(yob < 1900) |> 
  pull(speaker_anon) |>
  unique()

mid_speakers <- price |> 
  filter(between(yob, 1900, 1950)) |> 
  pull(speaker_anon) |>
  unique()

late_speakers <- price |> 
  filter(yob > 1950) |> 
  pull(speaker_anon) |>
  unique()

speakers_to_keep <- c(
  sample(early_speakers, size = 50),
  sample(mid_speakers, size = 50),
  sample(late_speakers, size = 50)
)

price_filtered <- price_filtered |> 
  filter(
    speaker_anon %in% speakers_to_keep,
    traj_no <= 3
  )
```

## GAMM with Random Smooths

```{r}
#| cache: true
#| code-line-numbers: "|8|10-11"
f1_fs_fit <- bam(
  formula = f1 ~ sex + 
    s(measurement_no, by = sex, k = 4, bs = "cr") +
    s(yob, by = sex, k = 4, bs = "cr") + 
    ti(measurement_no, yob, by = sex, k = c(4, 4), bs = "cr") +
    s(syllables_per_sec, k = 4, bs = "cr") + 
    s(speaker_anon, bs = "re") +
    s(measurement_no, traj_id, bs = "fs", k = 4, m = 1),
  data = price_filtered,
  discrete = TRUE,
  nthreads = detectCores() - 1
)
```

---

```{r}
# Note re.test = FALSE! This speeds things up
# a lot!
summary(f1_fs_fit, re.test = FALSE)
```

---

```{r}
appraise(f1_fs_fit)
```

---

```{r}
draw(f1_fs_fit)
```

---

```{r}
#| cache: true
#| output-location: slide
inspect_random(
  f1_fs_fit,
  select = 9,
  cond = list(
    "traj_id" = sample(unique(price_filtered$traj_id), size = 20)
  ),
  print.summary = FALSE
)
```


---

```{r}
# Here, the plots are getting a bit complicated.
big_mod_preds <- get_predictions(
  f1_fs_fit,
  cond = list(
    yob = c(1860, 1880, 1900, 1920, 1950, 1970),
    measurement_no = 1:11,
    sex = c("M", "F")
  )
)
```

---

```{r}
#| output-location: slide
big_mod_preds |> 
  ggplot(
    aes(
      x = measurement_no, 
      ymin = fit - CI,
      ymax = fit + CI,
      y = fit,
      fill = sex
    )
  ) +
  geom_line(aes(colour = sex)) +
  geom_ribbon(alpha = 0.2) +
  facet_wrap(vars(yob))
```

## Random Slopes

- For UC researchers (including research students), "Big R" (and François Bissey) is your friend.
    - 384GB of RAM
    - 32 cpus
    - 2TB of disk --- expandable to some extent to accommodate your projects 
- <https://wiki.canterbury.ac.nz/display/RCC/RStudio+instances+on+the+RCC>
- You don't need to lock up your computer over night!

## Next Time

- GAMMs are often used to investigate time series data.
- Time series data often suffers from [autocorrelation]{.red}:
    - i.e. the value at time $t$ is correlated with the value at time $t+1$.
- Random smooths can sometimes handle this (but are expensive)
- We'll look at testing for autocorrelation and other methods for handling it.
- We'll also look at [model comparison]{.red}

## What to do now?

::: {.fragment}

```{r}
#| eval: false
#install.packages('usethis')
usethis::use_course('nzilbb/gamm_sessions')
```

:::

- A new R project should open in R Studio.
- Go to the `session_2.R` script in the `scripts folder`.
- Work through the examples.

# References

```{r}
#| echo: false
grateful::nocite_references(
  grateful::cite_packages(output = "citekeys", out.dir = here())
)
```


::: refs

:::
